---
title: "Logistic Regression Project"
author: carmemias
date: 11th July 2020
output: html_notebook
---

# Project Aim

Use the UCI Adult data to predict individuals in the dataset earn more than $50K. The variables description can be found on the [UCI website](http://archive.ics.uci.edu/ml/datasets/Adult).

```{r install-packages, echo=FALSE}
for(p in c('Amelia', 'ggplot2', 'dplyr')) { 
  if(!(p %in% rownames(installed.packages))){
    install.packages(p)
  }
}

library(Amelia)
library(ggplot2)
library(dplyr)
```

```{r import-training_data}
vars <- c('age', 'workclass', 'fnlwgt', 'education', 'education-num', 'marital-status', 'occupation', 'relationship', 'race', 'sex', 'capital-gain', 'capital-loss', 'hours-per-week', 'native-country', 'earnings')
adult_training_data <- read.csv('data/adult.data', header = FALSE, sep = ",", quote = "", strip.white = TRUE, col.names = vars)
str(adult_training_data)
head(adult_training_data)
```

```{r import-test-data}
adult_test_data <- read.csv('data/adult.test', skip = 1, header = FALSE, sep = ",", quote = "", strip.white = TRUE, col.names = vars)
str(adult_test_data)
head(adult_test_data)
```

# Explore the training data

First, there are a lot of fields with a "?" value. Let's set them to NA and then find out how much missing data there is.

```{r set-NAs}
adult_training_data[ adult_training_data == "?" ] <- NA
```

```{r missing-data}
Amelia::missmap(adult_training_data, main = "Missing Map", col = c('yellow', 'black'), legend=FALSE)
```

We can see that 3 columns have missing data: occupation, workclass and native.country.

```{r how-much-missing-data}
num_empties <- function(cl) {
  num <- nrow(subset(adult_training_data, is.na(adult_training_data[cl])))
  perc <- round(num * 100 / nrow(adult_training_data), 2)
  
  list(num = num, perc = perc)
}

print("How much missing data:")
for(c in colnames(adult_training_data)){
  if(anyNA(adult_training_data[c])){
      print(paste("-", c, ":", num_empties(c)$num, "empty values. Or", num_empties(c)$perc, "% of the total.", sep = " " ))
  }
}

```
# Clean the data



