---
title: "ISLR_Chapter2"
author: "Carme Mias"
date: "02/05/2020"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
install.packages('https://cran.r-project.org/src/contrib/ISLR_1.2.tar.gz', repos=NULL, type='source')
require(ISLR)
```
# Applied Exercises Chapter 2.

## Ex.8 p. 54
The College dataset.

First, explore the data:
```{r import_college_data}
college_data <- read.csv('data/College.csv')
summary(college_data)
```

We'll turn the row names into a column called College.Name:
```{r college_row_names}
rownames(college_data)=college_data[,1]
colnames(college_data)[1] <- "College.Name"
head(college_data)
```

Let's have a look at the data in the first 10 columns:
```{r plot_first_10_columns}
pairs(college_data[,1:10])
```

And plot `Outstate` with `Private`:
```{r outstate_with_private}
invisible(attach(college_data))
plot(as.factor(Private), Outstate, ylab = "Outstate", xlab="Private")
```
The results show that private colleges take more out of state students than non-private colleges.

Add new column `Elite`, with "Yes"/"No" value to indicate whether at lest 50% of their students belong to the top 10% of their high school classes. Here's how it can be done:
1. We'll start by creating a vector "Elite" with as many "No" values as rows college_data has. 
2. Then we use college_data's `Top10perc` column as a filter to change the required vector values to "Yes".
3. We use the resulting vector to create a new column. The vector name will automatically be used as the column name.
```{r new_elite_column}
Elite <- rep("No", nrow(college_data))
Elite[college_data$Top10perc > 50] <- "Yes"
college_data <- data.frame(college_data, Elite)
head(college_data)
```

Let's see it in a graph:
```{r elite_plot}
plot(as.factor(Elite), Outstate, xlab="Elite", ylab="Outstate")
```
The results show that the elite colleges have far more out of state students than the non-elite colleges.

Finally, we'll explore some of the numerical variables.

```{r application_acceptance_enrollment_histograms}
op <- par(mfrow=c(1,3))
hist(Apps[Apps <= 30000], col = "lightblue", xlab = "Student Applications")
hist(Accept[Accept <=20000], col = "green", xlab = "Number of Acceptances")
hist(Enroll, col = "orange", xlab = "Number of Enrollments")
```

Let's see how private/public and elite/non-ellite colleges compare in terms of the data explored above:
```{r further_comparisons}

ordered_by_top10perc <- college_data[order(-Top10perc),]
attach(ordered_by_top10perc)
top_10_private_elite <- head(ordered_by_top10perc[Private=="Yes" & Elite=="Yes",], 10)
top_10_public_elite <- head(ordered_by_top10perc[Private=="No" & Elite=="Yes",], 10)

op <- par(mfrow=c(1,2))
barplot(cbind(top_10_private_elite$Apps, top_10_private_elite$Accept, top_10_private_elite$Enroll) ~ top_10_private_elite$College.Name, beside=TRUE, col=c("lightblue", "green", "orange"), legend.text = TRUE, xlab="College Name", main = "Top 10 Private Elite Colleges")

barplot(cbind(top_10_public_elite$Apps, top_10_public_elite$Accept, top_10_public_elite$Enroll) ~ top_10_public_elite$College.Name, beside=TRUE, col=c("lightblue", "green", "orange"), legend.text = TRUE, xlab="College Name", main = "Top 10 Public Elite Colleges")
```

## Ex.10 p. 55
The Boston dataset. The dataset does not include town names.

```{r import_ex10_data}
library(MASS)
str(Boston)
head(Boston)
```

We'll do some initial explorations:
```{r initial_boston_exploration}
pairs(Boston[,c("crim", "zn", "indus", "nox", "rm", "dis", "rad", "tax", "ptratio", "medv")])
```

Some interesting patterns seem to appear around Crime rates. Let's expore them further:
```{r crime_rate_comparisons}
attach(Boston)
op <- par(mfrow=c(2,3))
plot(x=zn, y=crim, xlab="Proportion of residential plots over 25,000 sq.ft", ylab="Capita Crime Rate")
plot(x=medv, y=crim, xlab="Median value of ower-occupied homes in $1000s", ylab="Capita Crime Rate")
plot(x=ptratio, y=crim, xlab="Pupil-teacher ratio", ylab="Capita Crime Rate")
plot(x=indus, y=crim, xlab="Proportion of non-retail business acres", ylab="Capita Crime Rate")
plot(x=dis, y=crim, xlab="Distance to employment centre", ylab="Capita Crime Rate")
plot(x=rad, y=crim, xlab="Accessibility to Radial Highway", ylab="Capita Crime Rate")
```

Also, some pollution and population level comparisons to re. distance to highway and work centre:
```{r work_population_and_pollution_comparisons}
op <- par(mfrow=c(3,2))
plot(x=rad, y=nox, xlab="Radial Highway accessibility rate", ylab="Nitrogen Oxides Concentration")
plot(x=indus, y=nox, xlab="Proportion of non-retail business acres", ylab="Nitrogen Oxides Concentration")
plot(x=dis, y=nox, xlab="Distance to employment centre", ylab="Nitrogen Oxides Concentration")
plot(x=indus, y=rad, xlab="Proportion of non-retail business acres", ylab="Radial Highway accessibility rate")
plot(x=lstat, y=nox, xlab="Lower status population (%)", ylab="Nitrogen Oxides Concentration")
plot(x=lstat, y=rm, xlab="Lower status population (%)", ylab="Rooms per dwelling")
```

Finally, a deeper look at the suburbs with the highest and lowest proportion of lowest status population:
```{r lowest_status_population}
highest_lstat <- max(Boston[,"lstat"])
invisible(nrow(Boston[lstat==highest_lstat,]))
suburbA <- Boston[lstat==highest_lstat,]

lowest_lstat <- min(Boston[,"lstat"])
invisible(nrow(Boston[lstat==lowest_lstat,]))
suburbB <- Boston[lstat==lowest_lstat,]
  
print(paste0("SuburbA has the highest percentage of lower-status population at: ", highest_lstat,"%" ))
print(paste0("SuburbB has lowest percentage of lower-status population at: ", lowest_lstat,"%" ))
```

Let's prepare the data from those two suburbs before we start:
```{r suburb_comparison_setup}
highest_and_lowest <- Boston[Boston$lstat==lowest_lstat | Boston$lstat==highest_lstat,]
invisible(highest_and_lowest[order(lstat),])
rownames(highest_and_lowest)<-c("suburbB", "suburbA")
pairs(highest_and_lowest)
```

And now, we can plot the various variables:
```{r suburb_comparison}
invisible(attach(highest_and_lowest))
# c("crim","indus", "nox", "rm", "age", "dis", "rad", "tax", "ptratio", "black", "medv")
barplot(cbind(crim, ptratio, rm, rad, dis) ~ lstat, beside=TRUE, names.arg = c("suburbB", "suburbA"), legend.text=c("Crime Rate", "Pupil-Teacher Ratio", "Average number of rooms", "Distance to highway", "Distance to work centre"), xlab = "Suburbs with Lowest and Highest proportion of lower-status population")
```

The variable relating to race seems to lack in detail but it is not shown to be a factor in the population status:
```{r race_vs_lstat}
barplot(black ~ lstat, ylab="Proportion of Black population", xlab = "Suburbs with Lowest and Highest proportion of lower-status population", names.arg = c("suburbB", "suburbA"))
```