---
title: "ISLR_Chapter2"
author: "Carme Mias"
date: "02/05/2020"
output: html_document
---

```{r setup, include=FALSE}
install.packages('https://cran.r-project.org/src/contrib/ISLR_1.2.tar.gz', repos=NULL, type='source')
require(ISLR)
```
# Applied Exercises Chapter 2.

## Ex.8 p. 54
The College dataset.

First, explore the data:
```{r import_college_data}
college_data <- read.csv('data/College.csv')
summary(college_data)
```

We'll turn the row names into a column called College.Name:
```{r college_row_names}
rownames(college_data)=college_data[,1]
colnames(college_data)[1] <- "College.Name"
head(college_data)
```

Let's have a look at the data in the first 10 columns:
```{r plot_first_10_columns}
pairs(college_data[,1:10])
```

And plot `Outstate` with `Private`:
```{r outstate_with_private}
attach(college_data)
plot(as.factor(Private), Outstate, ylab = "Outstate", xlab="Private")
```
The results show that private colleges take more out of state students than non-private colleges.

Add new column `Elite`, with "Yes"/"No" value to indicate whether at lest 50% of their students belong to the top 10% of their high school classes. Here's how it can be done:
1. We'll start by creating a vector "Elite" with as many "No" values as rows college_data has. 
2. Then we use college_data's `Top10perc` column as a filter to change the required vector values to "Yes".
3. We use the resulting vector to create a new column. The vector name will automatically be used as the column name.
```{r new_elite_column}
Elite <- rep("No", nrow(college_data))
Elite[college_data$Top10perc > 50] <- "Yes"
college_data <- data.frame(college_data, Elite)
head(college_data)
```

Let's see it in a graph:
```{r elite_plot}
plot(as.factor(Elite), Outstate, xlab="Elite", ylab="Outstate")
```
The results show that the elite colleges have far more out of state students than the non-elite colleges.

Finally, we'll explore some of the numerical variables.

```{r application_acceptance_enrollment_histograms}
op <- par(mfrow=c(1,3))
hist(Apps[Apps <= 30000], col = "lightblue", xlab = "Student Applications")
hist(Accept[Accept <=20000], col = "green", xlab = "Number of Acceptances")
hist(Enroll, col = "orange", xlab = "Number of Enrollments")
```

Let's see how private/public and elite/non-ellite colleges compare in terms of the data explored above:
```{r further_comparisons}

ordered_by_top10perc <- college_data[order(-Top10perc),]
attach(ordered_by_top10perc)
top_10_private_elite <- head(ordered_by_top10perc[Private=="Yes" & Elite=="Yes",], 10)
top_10_public_elite <- head(ordered_by_top10perc[Private=="No" & Elite=="Yes",], 10)

op <- par(mfrow=c(1,2))
barplot(cbind(top_10_private_elite$Apps, top_10_private_elite$Accept, top_10_private_elite$Enroll) ~ top_10_private_elite$College.Name, beside=TRUE, col=c("lightblue", "green", "orange"), legend.text = TRUE, xlab="College Name", main = "Top 10 Private Elite Colleges")

barplot(cbind(top_10_public_elite$Apps, top_10_public_elite$Accept, top_10_public_elite$Enroll) ~ top_10_public_elite$College.Name, beside=TRUE, col=c("lightblue", "green", "orange"), legend.text = TRUE, xlab="College Name", main = "Top 10 Public Elite Colleges")
```