---
title: "Bike Sharing Project"
author: Carme Mias
date: June 2020
output: html_notebook
---

Project exercise from the "Data Science and machine Learning bootcamp with R" course (section 21, Linear Regression) and using the data from the [Bike Sharing](https://www.kaggle.com/c/bike-sharing-demand/overview) Kaggle competition.

Data available to download from the Kaggle website.

```{r required-libraries, include=FALSE, eval=TRUE}
install.packages(c(ggplot2, corrplot, corrgram))
require('ggplot2')
require('corrplot')
require('corrgram')
```

## Get the Data

```{r initial-data-frames}
training_data <- read.csv('data/train.csv')
test_data <- read.csv('data/test.csv')

head(training_data)
summary(training_data)
nrow(training_data)
```

The variable we are trying to predict is `count`.

## Data Exploration

First, a temperature/count scatterplot:

```{r temperature-count-plot}
ggplot( training_data, aes( x=temp, y=count, color=temp ) ) + 
  geom_point( alpha=0.3 ) + scale_color_gradient(low = 'blue', high = 'orange')  +
  xlab( 'Temperature, Celsius' ) + ylab( 'Number of bikes rented' ) + ggtitle( 'Temperature/Count scatterplot' )
```

Then, a datetime/count scatterplot:

```{r datetime-count-scatterplot}
ggplot( training_data, aes( x=as.POSIXct(datetime), y=count, color=temp ) ) + 
  geom_point( alpha=0.3 ) + scale_color_gradient(low = 'blue', high = 'orange')  +
  xlab( 'Date and time' ) + ylab( 'Number of bikes rented' ) + ggtitle( 'Datetime/Count scatterplot' )
```
From the graph, we can conclude that, in general, the number of bicycles rented has been growing steadily and that there are more rentals when the temperature is high than when it's low.

## The Linear Regression Model

```{r temp-count-correlation}
temp_count_cor <- cor(training_data[,c('temp', 'count')])
print(temp_count_cor)
```

```{r season-count-boxplot}
ggplot(training_data, aes(x=as.factor(season), y=count, color = as.factor(season))) +
  geom_boxplot() + scale_x_discrete(labels=c("1"="Spring (1)", "2"="Summer (2)", "3"="Autumn (3)", "4"="Winter (4)")) +
  xlab( 'Season' ) + ylab( 'Number of bikes rented' ) + ggtitle( 'Season/Count boxplot' )
```

Two characteritics stand out: one is that there are a lot of outlying values, the other is that there seemed to have been more bicycles rented in winter than in spring.

## Feature Engineering

We'll separate the time from the date and have it in a separate column called *time*.

```{r time-column}
time <- as.POSIXlt(training_data[,'datetime'])$hour
training_data <- cbind(training_data, time)
head(training_data)
```

Now we can see the bike use variations according to time of the day:

```{r by-time-for-working-days}
ggplot(subset(training_data, workingday == 1), aes(x=time, y=count, color=temp)) + 
  geom_point(size = 2, alpha=0.3, position=position_jitter(w=0.35, h=0)) + scale_color_gradient(low = 'blue', high = 'orange')  +
  xlab( 'Time of Day' ) + ylab( 'Number of bikes rented' ) + ggtitle( 'Time of Day/Count on Working Days' )
```

```{r by-time-for-non-working-days}
ggplot(subset(training_data, workingday == 0), aes(x=time, y=count, color=temp)) + 
  geom_point(size = 2, alpha=0.3, position=position_jitter(w=0.35, h=0)) + scale_color_gradient(low = 'blue', high = 'orange')  +
  xlab( 'Time of Day' ) + ylab( 'Number of bikes rented' ) + ggtitle( 'Time of Day/Count on Non-Working Days' )
```

## Build the Model