---
title: "Bike Sharing Project"
author: Carme Mias
date: June 2020
output: html_notebook
---

Project exercise from the "Data Science and machine Learning bootcamp with R" course (section 21, Linear Regression) and using the data from the [Bike Sharing](https://www.kaggle.com/c/bike-sharing-demand/overview) Kaggle competition.

Data available to download from the Kaggle website.

```{r required-libraries, include=FALSE, eval=TRUE}
install.packages(c(ggplot2))
require('ggplot2')
```

## Get the Data

```{r initial-data-frames}
training_data <- read.csv('data/train.csv')
test_data <- read.csv('data/test.csv')

head(training_data)
summary(training_data)
nrow(training_data)
```

The variable we are trying to predict is `count`.

## Data Exploration

First, a temperature/count scatterplot:

```{r temperature-count-plot}
ggplot( training_data, aes( x=temp, y=count, color=temp ) ) + 
  geom_point( alpha=0.3 ) + scale_color_gradient(low = 'blue', high = 'orange')  +
  xlab( 'Temperature, Celsius' ) + ylab( 'Number of bikes rented' ) + ggtitle( 'Temperature/Count scatterplot' )
```

Then, a datetime/count scatterplot:

```{r datetime-count-scatterplot}
ggplot( training_data, aes( x=as.POSIXct(datetime), y=count, color=temp ) ) + 
  geom_point( alpha=0.3 ) + scale_color_gradient(low = 'blue', high = 'orange')  +
  xlab( 'Date and time' ) + ylab( 'Number of bikes rented' ) + ggtitle( 'Datetime/Count scatterplot' )
```

## The Linear Regression Model

